Properties and Bayes fitting of restricted Boltzmann machines

### 2017-07-08 writing/paper-rbm/appendices/appendix-center.Rmd
paper to bookdown

---
title: "Supplementary material to:  On the propriety of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
    number_sections: yes
    template: ../latex/jasa_article.tex
blind: false
---

Supplement material to follow consists of Section \ref{on-the-one-to-one-correspondence-between-parameters-and-moments-in-the-rbm-distribution}, providing proofs and discussion of results relating to the data coding in restricted Boltzmann machine models (RBMs).

# On the one-to-one correspondence between parameters and moments in the RBM distribution

For integers $m,n \geq 1$, consider random vectors $\bm{V}_m \equiv(V_1,\ldots,V_m)$ and $\bm{H}_n\equiv (H_1,\ldots,H_n)$, where the random variables $V_i$ and $H_j$ assume values in $\{1,-1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$. We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution, i.e. has probability mass function
\begin{align*}
&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta}) \propto \exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \\
& \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
\end{align*}
with $m+n+mn$ real parameters $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}$ and
$\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$ collected in a parameter vector denoted as $\bm{\theta}\in \mathbb{R}^{m+n+mn}$.
Denote the distribution of $(\bm{V}_m,\bm{H}_n)$ as RBM$(m,n,\bm{\theta})$. In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X=1)=1/2=P(X=-1)$.

@@

\begin{itemize}
\item The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below. Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
\item The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
\item The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).  This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.
\end{itemize}

\begin{apptheorem} 
\label{thm:one}
Let $m,n\geq 1$ and suppose the random vector   $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j  $ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
\item $\bm{\theta}_1 =\bm{\theta}_2$.
\item $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\
\end{enumerate}
\end{apptheorem}

**Proof of Theorem \ref{thm:one}.** We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
 claim 3 holds trivially. Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$. Let  $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}$, $\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the components of $\bm{\theta}_1$ and let $\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}$, $\tilde{\theta}_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the analogous components of $\bm{\theta}_2$. Then, by assumption,
\begin{align*}
\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{align*}
for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$. This result further implies $\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.  As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta_{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

@@

To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1. Hence, it suffices now to establish claim 3 from claim 1. This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).

\begin{applemma} 
\label{lemma:one}
Let $m,n\geq 1$.  Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$. Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are conditionally independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are conditionally independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.

Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$,
$$
P_1( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n), \qquad
$$
follows if $\E_1 V_i = \E_2 V_i$, $\E_1 H_j = \E_2 H_j$, $\E_1 V_i H_j = \E_2 V_j H_j$ holds for all $i\in\{1,\ldots,m\}$, $j\in \{1,\ldots,n\}$.
\end{applemma}

**Proof.** The case $m=n=1$ follows from Lemma \ref{lemma:two}. We now use an induction argument, assuming that the result of Lemma \ref{lemma:one} holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$. We shall treat the case $(m+1,n)$ (where the other case $(m,n+1)$ follows by symmetrical arguments).

Considering random variables $(\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)$, the induction hypothesis
applies to $(\bm{V}_{m},\bm{H}_n)$ where $\bm{V}_m =(V_1,\ldots,V_m)$. Hence, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$, we have $P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0$, $P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0$ and consequently
\begin{align}\label{eqn0}
P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
\end{align}
Likewise, by the induction hypothesis applied to $(\bm{V}_{m-},\bm{H}_n)$ for $\bm{V}_{m-} = (V_2,\ldots,V_{m+1})$, we have
\begin{eqnarray}\label{eqn1}
 1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
\nonumber  &=&  P_2(V_{m+1} = a | \bm{H}_n=\bm{h}_n)\\
\nonumber  & = &1-P_2(V_{m+1} = b | \bm{H}_n=\bm{h}_n)
\end{eqnarray}
for any $\bm{h}_n \in \{a,b\}^{n}$. Now fix  $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$ and let $\bm{v}_{m+1}= (v_{m+1},\bm{v}_m)$ for $v_{m+1}\in\{a,b\}$.  Then, by the conditional independence assumption,
$$
P_i( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_i( V_{m+1}=v_{m+1}| \bm{H}_n=\bm{h}_n) P_i( \bm{V}_{m}=v_{m}| \bm{H}_n=\bm{h}_n)
$$
for $i=1,2$,  so that we have $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\ref{eqn0})-(\ref{eqn1}). Consequently, it follows that
\begin{eqnarray*}
P_1( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n) &=& P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) P_1(\bm{H}_n=\bm{h}_n)\\
& =& P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)P_2(\bm{H}_n=\bm{h}_n) \\&=&P_2( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n)
\end{eqnarray*}
by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma \ref{lemma:one}. $\Box$

\begin{applemma} 
\label{lemma:two}
Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
under two probability distributions $P_1$ and $P_2$. Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\item $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
\item $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.
\end{enumerate}
\end{applemma}

**Proof.** Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$. The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
$$
(c_1+c_2) = (d_1+d_2), \quad  (c_1+c_3) = (d_1+d_3) \quad (b-a) c_1 + a (2c_1+c_2+c_3) =  (b-a) d_1 + a (2d_1+d_2+d_3).
$$
As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3)$ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$. Hence, claim 1 implies claim 2 in Lemma \ref{lemma:two}. Claim 2 also trivially implies claim 1. $\Box$

\begin{apptheorem} 
\label{thm:two}
Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.  Then, the following are equivalent
\begin{enumerate}
\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}$.
\item $V_1,\ldots,V_m, H_1,\ldots,H_n$ are iid Bernoulli$(1/2)$ random variables.\\
\end{enumerate}
\end{apptheorem}

**Proof of Theorem \ref{thm:two}.**  We first establish the equivalence of claims 2 and 3. If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.

@@

To establish the equivalence of claims 1 and 2 in Theorem \ref{thm:two} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1. By Theorem \ref{thm:one}, if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$. $\Box$

>>>>

\clearpage

# (APPENDIX) Appendix {-}

# On the one-to-one correspondence between parameters and moments in the RBM distribution {#appendix-rbm}

For integers $m,n \geq 1$, consider random vectors $\bm{V}_m \equiv(V_1,\ldots,V_m)$ and $\bm{H}_n\equiv (H_1,\ldots,H_n)$, where the random variables $V_i$ and $H_j$ assume values in $\{1,-1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$. We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution, i.e. has probability mass function
\begin{align*}
&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta}) \propto \exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \\
& \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
\end{align*}
with $m+n+mn$ real parameters $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}$ and
$\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$ collected in a parameter vector denoted as $\bm{\theta}\in \mathbb{R}^{m+n+mn}$.
Denote the distribution of $(\bm{V}_m,\bm{H}_n)$ as RBM$(m,n,\bm{\theta})$. In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X = 1) = 1/2 = P(X = -1)$.

@@

- The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \@ref(thm:one) below. Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
- The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
- The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).  This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \@ref(thm:two).

```{theorem, label="one", echo=TRUE}
Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent

1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta}_1 =\bm{\theta}_2$.
3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\

```

```{proof, echo=TRUE}
We first establish the equivalence of claims 2 and 3. If claim 2 holds, then claim 3 holds trivially. Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$. Let  $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}$, $\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the components of $\bm{\theta}_1$ and let $\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}$, $\tilde{\theta}_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the analogous components of $\bm{\theta}_2$. Then, by assumption,
\begin{align*}
\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{align*}
for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$. This result further implies $\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.  As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta_{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

@@

To establish the equivalence of claims 1 and 3 in Theorem \@ref(thm:one) under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1. Hence, it suffices now to establish claim 3 from claim 1. This follows immediately from Lemma \@ref(lem:first) below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
```

```{lemma, label="first", echo=TRUE}
Let $m,n\geq 1$. Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$. Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are conditionally independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are conditionally independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.

Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$,
$$
P_1( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n), \qquad
$$
follows if $\E_1 V_i = \E_2 V_i$, $\E_1 H_j = \E_2 H_j$, $\E_1 V_i H_j = \E_2 V_j H_j$ holds for all $i\in\{1,\ldots,m\}$, $j\in \{1,\ldots,n\}$.
```

```{proof, echo=TRUE}
The case $m=n=1$ follows from Lemma \@ref(lem:second). We now use an induction argument, assuming that the result of Lemma \@ref(lem:first) holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$. We shall treat the case $(m+1,n)$ (where the other case $(m,n+1)$ follows by symmetrical arguments).

Considering random variables $(\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)$, the induction hypothesis applies to $(\bm{V}_{m},\bm{H}_n)$ where $\bm{V}_m =(V_1,\ldots,V_m)$. Hence, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$, we have $P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0$, $P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0$ and consequently
\begin{align}(\#eq:0)
P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
\end{align}
Likewise, by the induction hypothesis applied to $(\bm{V}_{m-},\bm{H}_n)$ for $\bm{V}_{m-} = (V_2,\ldots,V_{m+1})$, we have
\begin{eqnarray}(\#eq:0)
 1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
\nonumber  &=&  P_2(V_{m+1} = a | \bm{H}_n=\bm{h}_n)\\
\nonumber  & = &1-P_2(V_{m+1} = b | \bm{H}_n=\bm{h}_n)
\end{eqnarray}
for any $\bm{h}_n \in \{a,b\}^{n}$. Now fix  $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$ and let $\bm{v}_{m+1}= (v_{m+1},\bm{v}_m)$ for $v_{m+1}\in\{a,b\}$.  Then, by the conditional independence assumption,
$$
P_i( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_i( V_{m+1}=v_{m+1}| \bm{H}_n=\bm{h}_n) P_i( \bm{V}_{m}=v_{m}| \bm{H}_n=\bm{h}_n)
$$
for $i=1,2$,  so that we have $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\@ref{eq:0})-(\@ref(eq:1)). Consequently, it follows that
\begin{eqnarray*}
P_1( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n) &=& P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) P_1(\bm{H}_n=\bm{h}_n)\\
& =& P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)P_2(\bm{H}_n=\bm{h}_n) \\&=&P_2( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n)
\end{eqnarray*}
by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma \@ref(lem:first). 
```

```{lemma, label="second", echo=TRUE}
Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
under two probability distributions $P_1$ and $P_2$. Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, the following are equivalent

1. $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
2. $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.

```

```{proof, echo=TRUE}
Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$. The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
$$
(c_1+c_2) = (d_1+d_2), \quad  (c_1+c_3) = (d_1+d_3) \quad (b-a) c_1 + a (2c_1+c_2+c_3) =  (b-a) d_1 + a (2d_1+d_2+d_3).
$$
As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3)$ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$. Hence, claim 1 implies claim 2 in Lemma \@ref(lem:second). Claim 2 also trivially implies claim 1. 
```

```{theorem, label="two", echo=TRUE}
Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.  Then, the following are equivalent

1. $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}$.
3. $V_1,\ldots,V_m, H_1,\ldots,H_n$ are iid Bernoulli$(1/2)$ random variables.

```

```{proof, echo=TRUE}
We first establish the equivalence of claims 2 and 3. If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.

@@

To establish the equivalence of claims 1 and 2 in Theorem \@ref(thm:two) under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1. By Theorem \@ref(thm:one), if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$. 
```

# Flexibility of the RBM model {#appendix-rbm2}

In an RBM model with enough hidden variables, parameter values may be chosen to match any given cell probabilities with arbitrary closeness. Additionally, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small. We show this for a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.
